---
title: "cranrepo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cranrepo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Set up a CRAN-like repository

cranrepo gives you the basic tools you need to set up a CRAN-like repository.
In this vignette, we'll go through the steps you'd take to do that.

```{r setup}
library(cranrepo)
```

## Create a repository directory

The first step to creating a repository is to set up a directory structure,
and the `PACKAGES` files that act as the package indexes. cranrepo has a single
function to do both: `repo_create()`.

Let's set up a repository structure in a temporary location:

```{r create}
repo <- repo_create(fs::path_temp("cranrepo/repos/latest"))
```

The created directory structure contains distribution trees for all three main
types of packages: source, Windows binary, and Mac binary packages. That ensures
that our repository can be used via the `file:` protocol without errors on any
major platform without throwing errors about missing files; something that's of
particular usefulness when setting up an internal repository shared via a
network drive, for example.

```{r view-repo}
fs::dir_tree(repo)
```

In addition to the usual `PACKAGES` files, you'll notice our distribution trees
contain a `PACKAGES.db` file. This is the SQLite databse used by
[cranlike](https://cran.r-project.org/package=cranlike) under the hood
to manage the repository index in an efficient manner.

## Add packages to your repository

The next thing we'll need to do is to add some packages to our repository.
Let's get a package to insert from CRAN. We'll get both the source and binary
bundle. It's always good practice to serve the source bundle from your repository,
and the pre-built binary makes the package accessible to a wider audience.

```{r get-package}
op <- options(repos = c(CRAN = "https://cloud.r-project.org"))
src <- download.packages("ggplot2", tempdir(), type = "source")[, 2]
bin <- download.packages("ggplot2", tempdir(), type = "win.binary")[, 2]
options(op)
```

Instead of getting package bundles from another repository like CRAN, we could
just as well build them for our own local package ourselves with `R CMD build`
and `R CMD INSTALL --build`. All that matters is that we have the appropriate
bundle(s) available for inserting into our repository.

Now that we have some package bundles, we can add them to our repository with
`repo_insert()`. You will, however, need to let cranrepo know what type of
bundle you are inserting. That's one of the deliberate design decisions made
in an effort find the right level of abstraction for the package.

```{r insert}
repo_insert(repo, src, "source")
repo_insert(repo, bin, "win.binary")
```

We can verify that the packages were successfully inserted by checking the
available packages in our repo:

```{r available}
available_packages <- function(repos, type = "source") {
  available.packages(repos = repos, type = type)[, c("Package", "Version")]
}

available_packages(paste0("file:", repo), type = "source")
available_packages(paste0("file:", repo), type = "binary")
```

Congratulations! You've set up your own CRAN-like repository serving source and
binary packages. That said, there's still some operations you'll need in order
to manage the lifecycle of a package repository.

## Serving your repository over HTTP

Earlier, we accessed our repository from our local computer using the `file:`
protocol. Unless your repository was created on a network location, chances are
that other users won't be able to access the repository this way. Instead, you'll
need to either move the repository to a network location, or serve the files
over a web protocol like HTTP. cranrepo provides a convenience function,
`repo_serve()`, which can be used to start a simple HTTP server on your PC to
let others access our repository.

```r
repo_serve(repo)
```

By default, the repository is served on port 4321. We can check the available
packages again, from another R session:

```r
available_packages("http://localhost:4321")
```

`repo_serve()` is mainly provided for convenience and testing purposes when
developing a repository. When actually publishing a repository for you consumers,
it's like you'd have a different solution for hosting the repository as e.g. static files.

You now have the basics to set up a repository, add packages, and distribute
them to collaborators. Next, let's tackle a few admin tasks.

## Removing a package

Sometimes (hopefully rarely) you'll realize you added a package to your repository
that you really shouldn't have. To recover and remove the package, `repo_remove()`
comes to your aid. Let's remove all the versions of ggplot2 Windows binaries
from our repository, including the one we added earlier:

```{r remove}
removed <- repo_remove(repo, "ggplot2", version = NULL, type = "win.binary")
```

For transparency, we get back the paths to the files that were removed:

```{r removed}
removed
```

## Updating the package index

If you make external changes to the package repository, you may end up with a
situation where the package index is out of sync with the actual files in the
repository. For example, let's manually delete the source bundle of ggplot2 we
added earlier:

```{r remove-manual}
fs::file_delete(fs::path(repo, "src/contrib", fs::path_file(src)))
```

The repository still reports the package as being available:

```{r update-before}
available_packages(paste0("file:", repo))
```

We'll need to call `repo_update()` to bring the package index up to date,
and report to our consumers that the package is no longer available:

```{r update}
repo_update(repo, "source")
available_packages(paste0("file:", repo))
```

## Conclusion

To wrap up, let's clean up the files we created for this vignette:

```{r cleanup}
fs::dir_delete(repo)
```

There we go! You now have the basic tools and knowledge to set up and manage
your own CRAN-like repository. Use them wisely.
